From physical principles to ???

### 2016-02-04 2015CompleteDerivation/completeDerivation.tex
Rewrite of the fundamental model

We start by fixing a \emph{physical system} we want to study, that is something we can interact with and perform measurements on (e.g. a planet, water). We call \emph{environment} everything else. We set what particular aspect we want to study (e.g. the motion around a star, the flow in a pipe). We call \emph{state} a physically distinguishable configuration of the portion under study at a particular time. Since the state does not, in general, exhaust the description of the system, a part remains \emph{unstated}, and as such we'll call it, for lack of a better word. Note that the environment plays an essential role here as it's what allows us to define two states as physically distinguishable: we can find an external process (i.e. a different environment) whose outcome changes depending on the different state. As such processes are what we can use to perform measurements, we consider the experimental apparatus (and us performing measurements) independent of the system, part of the environment. This is true even in the case of general relativity, where we can imagine multiple researchers on small spaceships collecting data without greatly influencing the motion of stars and planets. We call this triad of state, unstated part and environment the \emph{fundamental model of physics}, as it is the simplest model that still captures the crucial aspects of studying and writing laws for a physical system.\footnote{The case where the physical system is the whole universe and there is no environment presents practical problems (e.g. What physical device can we use to store and process the state of the whole universe to make predictions and compare?) and conceptual challenges (e.g. How do we define physically distinguishable? Can the physical laws predict which measurements are we going to make? Does that limit what is actually distinguishable?) that the current physical theories do not seem to be equipped to address, and therefore will be absent from our discussion.} With this in mind, we introduce the following:

\begin{assump}[Determinism and reversibility]
The state of the physical system under study undergoes deterministic and reversible evolution.
\end{assump}

As the terms used have different meaning in different contexts, let's first clarify them. By deterministic (and reversible) we mean that future (and past) states are uniquely identified by the present state. We can then think of the state of the system as the part that interacts deterministically and reversibly with the environment (and itself) while the unstated part as interacting non-deterministically with the environment (and itself).

Much of the focus later will be given to the state under deterministic and reversible evolution. However, the non-deterministic part plays a fundamental role as well. In fact: it is this aspect of the evolution that determines what states are available and their description. Suppose we study the motion of a cannonball, its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. While light and air molecules scatter off its surface randomly, its trajectory is not greatly affect as it is a massive rigid body. Suppose we study the motion of a small particle, small enough that the random scattering does influence the trajectory and it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states have. The issue is that the set of states must be closed under the non-deterministic evolution as well. That is: if we apply the non-deterministic evolution to any state, we still need to end up in one of the possible states. If the Brownian motion is strong enough, the state gets knocked out of the set of states determined by a simple position/momentum pair.

@@

As we have seen, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is: given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the trajectories of the parts are not (those fall within the unstated part). In fact: we \emph{cannot} assume trajectories and states are always defined at all for the unstated part. Consider a muon and its decay into an electron and two neutrinos: it is clear that the three outgoing particles have a state and trajectory of their own, it is clear that the resulting total mass and energy came from the muon. Yet, before the decay, we cannot ascribe an internal independent state and evolution to each part. The state of a muon is not some combination of the state of an electron and two neutrinos. That is: the unstated part is not just microstates.

Now that we have clarified what we mean by deterministic and reversible evolution (even on a statistical ensemble), and how also the non-deterministic part is key to define states (as the set as a whole needs to be invariant under it), we can ask ourselves: when is this assumption valid?

\begin{rationale}
The claim is that deterministic and reversible evolution is necessary to be able to study a physical system. That is: the only part of a system that can be studied through reliable and reproducible experimentation or observation is the part that undergoes deterministic and reversible evolution. We can provide different arguments that point in the same direction.

First, to be able to identify the system, we must first be able to tell it apart from the environment. Intuitively, we can distinguish between two chairs because we can move the first to another room and sit on it without having touched the second. We can manipulate the state of the first system without affecting the second, and vice-versa. So, to tell a system apart from its environment, there has to to be a certain amount of isolation between the two: if I affect its state, the state of other systems is not affected and vice-versa. This means that the system future and past states are only determined by its own state and the state undergoes deterministic and reversible evolution.

Second, the aim of physics is to write laws that can be used to make prediction that can be validated experimentally. If I drop an anvil from a tower, it will accelerate at $9.81 m/s^2$; if I want the anvil to reach the ground at $x m/s$ I have to drop it from $y m$. To the extent that we want to make predictions in time, we need to have a one-to-one correspondence between initial and final states.

@@

This link between state definition and deterministic processes should not be too surprising as the state, in the context of thermodynamics and systems theory, is often defined as \emph{the set of variables needed to determine the future evolution of the system}. As we saw before, this applies also to statistical processes: the distribution (the ensemble) as a whole can be indeed calculated, measured and prepared. We can also describe the evolution of each element of the distribution provided that: we have a way to isolate it and study it under deterministic and reversible motion (so that we can define its states); the non-deterministic motion does not alter the system (the set of states remains the same). For example, consider a wooden ball described by position and momentum of the center of mass: if randomly scattered by other balls retain the set of states, if thrown into a fire it does not.

So, if it seems we need to require deterministic and reversible motion, why we call this an assumption? First of all, because it's an idealization: it can never be completely achieved in practice. A system can be prepared or measured up to a certain level of precision. Perfect isolation of a system is impossible both practically (e.g. black-body radiation, gravity, ...) and conceptually (e.g. if the system is perfectly isolated, we cannot interact with it: how can it studied through experimentation?). It's a simplifying assumption that can only be taken if the environment and the internal dynamics of the system interact in such a way that they little affect and are little affected by the aspect we are studying. As we saw before, for example, assuming that the states consists of the position and momentum of the center of mass requires assuming that the Brownian motion of the body is negligible.

The second reason is that technically the assumption comes first. If we start looking for a law, we are implicitly assuming that there exists one to be found. The opposite assumption, there is no law to be found, can't really be experimentally proven exhaustively (i.e. you can only show that you have not found it among a certain class). Very often the hard part in science is finding for what system and in what conditions such assumption hold. Finding the law after that is a much simpler endeavour, and is something that is routinely done by students in their intro labs.
\end{rationale}

>>>>

We start by fixing a \emph{physical system}, meaning something we can interact with and perform measurements on (e.g. a planet, a fluid, ...). We call \emph{environment} everything else. We set what particular aspect we want to study (e.g. the motion around a star, the flow in a pipe, ...). We call \emph{state} a physically distinguishable configuration of the aspect under study at a particular time (e.g. position/momentum of center of mass, velocity field, ...). Since the state does not, in general, exhaust the description of the system, a part remains \emph{unstated}, and as such we'll call it, for lack of a better word (e.g. the chemical composition, the motion of each of its molecules, ...). Note that the environment plays an essential role here as it's what allows us to define two states as physically distinguishable: we can find an external process (i.e. part of the environment) whose outcome changes depending on the different state. As such processes are what we can use to perform measurements, we consider the experimental apparatus (and us performing measurements) independent of the system, part of the environment.\footnote{This is true even in the case of general relativity, where we can imagine multiple researchers on small spaceships collecting data without greatly influencing the motion of stars and planets. The case where the physical system is the whole universe and there is no environment presents practical problems  and conceptual challenges that the current physical theories do not seem to be equipped to address, and therefore will be absent from our discussion. For example, what physical device can we use to store and process the state of the whole universe to make predictions and compare? How do we define physically distinguishable? Do the physical laws determine which measurements we are going to make and does that limit what is actually distinguishable?}

In this context, we define a process \emph{deterministic} if the state at a given time uniquely identifies states at future times, and \emph{reversible} if it identifies states at past times. This means that, in general, the evolution of the unstated part is non-deterministic and/or non-reversible in the sense that the state of the system is not enough to determine its evolution. For example, the overall motion of a fluid in a pipe can be considered deterministic even if the motion of each molecule is not. The motion is deterministic and/or reversible only as far as the state is concerned. With this in mind, we introduce the following:

\begin{assump}[Determinism and reversibility]\label{detrevass}
The state of the physical system under study undergoes deterministic and reversible evolution.
\end{assump}

\begin{rationale}
As it is an assumption, we first need to discuss when it is valid. More specifically, we need to understand that the non-deterministic/non-reversible evolution of the unstated part plays as much as a fundamental role as the deterministic/reversible evolution of the state. In fact: the non-deterministic part contributes in determining what states are available to the system.

Suppose we study the motion of a cannonball, its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. While light and air molecules scatter off its surface randomly, its trajectory is not greatly affect as it is a massive rigid body. Suppose we study the motion of a small particle, small enough that the random scattering does influence the trajectory and it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states have. The issue is that the non-deterministic evolution must not change the set of states itself. That is: if we apply the non-deterministic evolution to any state, we still need to end up in one of the possible states. If the Brownian motion is strong enough, a simple position/momentum pair is not sufficient.

@@

As we have hinted, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is: given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the trajectories of the parts are not (those fall within the unstated part). In fact: we \emph{cannot} assume trajectories and states are always defined at all for the unstated part. Consider a muon and its decay into an electron and two neutrinos: it is clear that the three outgoing particles have a state and trajectory of their own, it is clear that the resulting total mass and energy came from the muon. Yet, before the decay, we cannot ascribe an internal independent state and evolution to each part. The state of a muon is not some combination of the state of an electron and two neutrinos. That is: the unstated part is not just microstates.

It should be also clear that what constitute state and unstated part does not depend only on the system under study, but also on the processes we are considering. In some circumstances, the chemical composition of a fluid may be relevant, in others it may not. The choices of environment, state and unstated part are not independent from each other. By choosing a particular set of states, we are not only saying that the state evolution is well approximated by a deterministic/reversible map from initial to final state, but we are also saying that the non-deterministic/non-reversible evolution of the unstated part does not change the nature of the system, and processes that do not satisfy these conditions are not under consideration.

As with all assumptions, we should also ask whether it is necessary. That is: could we define a set of physically distinguishable states and yet have no deterministic and reversible processes defined on them? The claim is that this assumption is indeed needed, as without it we cannot properly define states or write useful physics laws. We can provide different arguments that point in the same direction.

First, to be able to identify the system, we must be able to tell it apart from anything else. Intuitively, we can distinguish between two chairs because we can move the first to another room and sit on it without having touched the second. We can manipulate the state of the first system without affecting the second, and vice-versa. So, to identify a system it has to be sufficiently isolated from everything else. This means that the system future and past states are with good approximation determined only by its own state: the state undergoes deterministic and reversible evolution.

Second, the aim of physics is to write laws that can be used to make predictions that can be validated experimentally. If I drop an anvil from a tower, it will accelerate at $9.81$~m/s$^2$; if I want the anvil to reach the ground at $x$~m/s I have to drop it from $y$~m. To the extent that we want to make predictions in time, we need to have a one-to-one correspondence between initial and final states.

@@

Fourth, to be able to ascribe a property to a system we need to claim that, at least for a finite period of time, the system either held or did not hold such property. That is: there is a deterministic and reversible process for that finite period of time for such property.

This link between state definition and deterministic processes should not be too surprising as the state, in the context of thermodynamics and systems theory, is often defined as \emph{the set of variables needed to determine the future evolution of the system}. As we saw before, this applies also to statistical processes: the distribution (the ensemble) as a whole can be indeed calculated, measured and prepared. We can also describe the evolution of each element of the distribution provided that: we have a way to isolate it and study it under deterministic and reversible motion (so that we can define microstates); the non-deterministic motion does not alter the system (the set of microstates is preserved by the evolution).

As with all assumptions, we should stress that it's an idealization: it can never be completely achieved in practice. A system can be prepared or measured up to a certain level of precision. Perfect determinism and isolation of a system is impossible both practically (e.g. black-body radiation, gravity, ...) and conceptually (e.g. if the system is perfectly isolated, we cannot interact with it: how can it be physically distinguished?). It's a simplifying assumption that can only be taken if the environment and the internal dynamics of the system interact in such a way that they little affect and are little affected by the aspect we are studying. As we saw before, for example, assuming that the state consists of the position and momentum of the center of mass requires assuming that the Brownian motion of the body is negligible.

Yet, this is a fundamental assumption in the sense that it is needed. If a particular set of states does not satisfy deterministic and reversible evolution under certain conditions, what we do is to keep at it until we find a set that does. That is: we work to restore the assumption. Finding new sets of states with new laws of evolution is, in fact, what leads to new physics. We will therefore call \emph{fundamental model of physics} the triad of state, unstated part and environment, with the assumption that the state may undergo deterministic and reversible evolution, and the unstated part undergoes non-deterministic non-reversible evolution that does not alter the set of states.
\end{rationale}

\section{States}
